---
title: "Pharmaceutical Dispensing Analysis"
output: html_document
---

```{r setup}
library(tidyverse)
library(sf)
library(dplyr)
library(ggplot2)
```

## Load the main pharmaceutical data

```{r}
pharms <- read_csv(file.choose())

head(pharms)
glimpse(pharms)
summary(pharms)
```

## Check missing values

```{r}
colSums(is.na(pharms))
```

## Check number of rows and duplicated rows

```{r}
nrow(pharms)
nrow(distinct(pharms))
```

## Convert NumDisps to numeric

Values reported as "<6" do not provide an exact number.
They are treated as missing for numerical analysis.

```{r}
pharms <- pharms %>%
  mutate(
    NumDisps_num = if_else(
      NumDisps == "<6",
      NA_real_,
      as.numeric(NumDisps)
    )
  )
```

## Number of "<6" values

```{r}
sum(pharms$NumDisps == "<6", na.rm = TRUE)
```

## Missing values after conversion

```{r}
sum(is.na(pharms$NumDisps_num))
```

## Identify the Top 10 Therapeutic Groups

```{r}
top10 <- pharms %>%
  group_by(TherapeuticGrp2) %>%
  summarise(
    total_dispensings = sum(
      NumDisps_num,
      na.rm = TRUE
    ),
    .groups = "drop"
  ) %>%
  arrange(desc(total_dispensings)) %>%
  slice_head(n = 10)

top10
```

## Figure 1: Top 10 Therapeutic Groups

```{r}
fig1 <- ggplot(
  top10,
  aes(
    x = reorder(
      TherapeuticGrp2,
      total_dispensings
    ),
    y = total_dispensings
  )
) +
  geom_col() +
  coord_flip() +
  scale_y_continuous(
    labels = scales::comma
  ) +
  labs(
    title = "Top 10 Therapeutic Groups by Initial Dispensings",
    x = "Therapeutic Group Level 2",
    y = "Total initial dispensings (2021–2025)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(
      size = 14,
      face = "bold"
    ),
    axis.text.y = element_text(
      size = 9
    )
  )

fig1

ggsave(
  "top10_groups.png",
  fig1,
  width = 9,
  height = 6,
  dpi = 300
)
```

## Figure 2: Trends in the Top 10 Therapeutic Groups

```{r}
trend_top10 <- pharms %>%
  filter(
    TherapeuticGrp2 %in% top10$TherapeuticGrp2
  ) %>%
  group_by(
    TherapeuticGrp2,
    YearDisp
  ) %>%
  summarise(
    total_dispensings = sum(
      NumDisps_num,
      na.rm = TRUE
    ),
    .groups = "drop"
  )

trend_top10
```

```{r}
fig2 <- ggplot(
  trend_top10,
  aes(
    x = YearDisp,
    y = total_dispensings,
    group = TherapeuticGrp2,
    colour = TherapeuticGrp2
  )
) +
  geom_line(
    linewidth = 1
  ) +
  geom_point(
    size = 2
  ) +
  scale_y_continuous(
    labels = scales::comma
  ) +
  scale_x_continuous(
    breaks = 2021:2025
  ) +
  labs(
    title = "Trends in Initial Dispensings for the Top 10 Therapeutic Groups",
    x = "Year",
    y = "Total initial dispensings",
    colour = "Therapeutic Group Level 2"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(
      size = 14,
      face = "bold"
    ),
    legend.position = "right"
  )

fig2

ggsave(
  "trend_top10.png",
  fig2,
  width = 9,
  height = 6,
  dpi = 300
)
```

## Load district-level data

```{r}
district_data <- read_csv(file.choose())

head(district_data)
glimpse(district_data)

nrow(district_data)

unique(district_data$District)
```

## Convert NumDisps and NumPpl to numeric

```{r}
district_data <- district_data %>%
  mutate(
    NumDisps_num = if_else(
      NumDisps == "<6",
      NA_real_,
      as.numeric(NumDisps)
    ),
    NumPpl_num = if_else(
      NumPpl == "<6",
      NA_real_,
      as.numeric(NumPpl)
    )
  )
```

## Remove New Zealand total

```{r}
district_data <- district_data %>%
  filter(
    District != "New Zealand"
  )

n_distinct(district_data$District)

unique(district_data$District)
```

## Figure 3: Total Initial Dispensings by District

```{r}
district_total <- district_data %>%
  group_by(District) %>%
  summarise(
    total_dispensings = sum(
      NumDisps_num,
      na.rm = TRUE
    ),
    .groups = "drop"
  ) %>%
  arrange(desc(total_dispensings))

district_total
```

```{r}
fig3 <- ggplot(
  district_total,
  aes(
    x = reorder(
      District,
      total_dispensings
    ),
    y = total_dispensings
  )
) +
  geom_col() +
  coord_flip() +
  scale_y_continuous(
    labels = scales::comma
  ) +
  labs(
    title = "Total Initial Dispensings by District",
    x = "District",
    y = "Total initial dispensings (2021–2025)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(
      size = 14,
      face = "bold"
    ),
    axis.text.y = element_text(
      size = 9
    )
  )

fig3

ggsave(
  "district_total.png",
  fig3,
  width = 9,
  height = 6,
  dpi = 300
)
```

## Top 10 Therapeutic Groups by District

```{r}
district_top10 <- district_data %>%
  filter(
    TherapeuticGrp2 %in% top10$TherapeuticGrp2
  ) %>%
  group_by(
    District,
    TherapeuticGrp2
  ) %>%
  summarise(
    total_dispensings = sum(
      NumDisps_num,
      na.rm = TRUE
    ),
    .groups = "drop"
  )

nrow(district_top10)

district_top10
```

## Create district order

```{r}
district_order <- district_total %>%
  arrange(total_dispensings) %>%
  pull(District)

district_top10 <- district_top10 %>%
  mutate(
    District = factor(
      District,
      levels = district_order
    ),
    TherapeuticGrp2 = factor(
      TherapeuticGrp2,
      levels = rev(top10$TherapeuticGrp2)
    )
  )
```

## Figure 4: Top 10 Therapeutic Groups by District

```{r}
fig4 <- ggplot(
  district_top10,
  aes(
    x = TherapeuticGrp2,
    y = District,
    fill = total_dispensings
  )
) +
  geom_tile() +
  scale_fill_continuous(
    labels = scales::comma
  ) +
  labs(
    title = "Top 10 Therapeutic Groups by District",
    x = "Therapeutic Group Level 2",
    y = "District",
    fill = "Total initial dispensings"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(
      size = 14,
      face = "bold"
    ),
    axis.text.x = element_text(
      angle = 45,
      hjust = 1,
      size = 8
    ),
    axis.text.y = element_text(
      size = 9
    )
  )

fig4

ggsave(
  "district_top10.png",
  fig4,
  width = 9,
  height = 7,
  dpi = 300
)
```

## Load district boundary shapefile

```{r}
dhb_shp <- st_read(file.choose())

names(dhb_shp)
```
## Drop geometry data for clarity, select district name and ID columns

```{r}
dhb_shp %>%
  st_drop_geometry() %>%
  select(starts_with("HD_L2")) %>%
  distinct()
```

## Check district name matching

```{r}
setdiff(district_total$District, dhb_shp$HD_L2_20_2)
setdiff(dhb_shp$HD_L2_20_2, district_total$District)
```

## Resolve district name mismatch

```{r}
dhb_shp <- dhb_shp %>%
  mutate(HD_L2_20_2 = if_else(
    HD_L2_20_2 == "Capital and Coast",
    "Capital & Coast",
    HD_L2_20_2
  ))
```

```{r}
setdiff(district_total$District, dhb_shp$HD_L2_20_2)
setdiff(dhb_shp$HD_L2_20_2, district_total$District)
```

## Join dispensing totals onto district geometry

```{r}
map_data <- dhb_shp %>%
  left_join(district_total, by = c("HD_L2_20_2" = "District"))
```

```{r}
# does each district appear once per row after join, or did the join fan out?
map_data %>%
  st_drop_geometry() %>%
  count(HD_L2_20_2) %>%
  arrange(desc(n))
```

```{r}
map_data_clean <- map_data %>%
  filter(HD_L2_20_2 != "Area outside health district" | is.na(HD_L2_20_2))
```

## Figure 5: Total Initial Dispensings (NumDisps) by District (Map)

```{r}
fig5 <- ggplot(map_data_clean) +
  geom_sf(
    aes(fill = total_dispensings),
    colour = "grey40",
    linewidth = 0.2
  ) +
  scale_fill_viridis_c(
    option = "plasma",
    labels = scales::comma,
    na.value = "grey90"
  ) +
  labs(
    title = "Total Initial Dispensings (NumDisps) by Health District (2021–2025)",
    fill = "Total initial\ndispensings"
  ) +
  theme_void() +
  theme(
    plot.title = element_text(
      size = 14,
      face = "bold"
    )
  )

fig5

ggsave(
  "district_total_map.png",
  fig5,
  width = 8,
  height = 8,
  dpi = 300
)
```

## Total Distinct Patients (NumPpl) by District

```{r}
district_total_ppl <- district_data %>%
  group_by(District) %>%
  summarise(
    total_ppl = sum(
      NumPpl_num,
      na.rm = TRUE
    ),
    .groups = "drop"
  ) %>%
  arrange(desc(total_ppl))

district_total_ppl
```

## Join patient totals onto district geometry

```{r}
map_data_ppl <- dhb_shp %>%
  left_join(district_total_ppl, by = c("HD_L2_20_2" = "District"))
```

```{r}
map_data_ppl %>%
  st_drop_geometry() %>%
  count(HD_L2_20_2) %>%
  arrange(desc(n))
```

```{r}
map_data_ppl_clean <- map_data_ppl %>%
  filter(HD_L2_20_2 != "Area outside health district" | is.na(HD_L2_20_2))
```

## Figure 6: Total Distinct Patients (NumPpl) by District (Map)

```{r}
fig6 <- ggplot(map_data_ppl_clean) +
  geom_sf(
    aes(fill = total_ppl),
    colour = "grey40",
    linewidth = 0.2
  ) +
  scale_fill_viridis_c(
    option = "plasma",
    labels = scales::comma,
    na.value = "grey90"
  ) +
  labs(
    title = "Total Distinct Patients (NumPpl) by Health District (2021–2025)",
    fill = "Total distinct\npatients"
  ) +
  theme_void() +
  theme(
    plot.title = element_text(
      size = 14,
      face = "bold"
    )
  )

fig6

ggsave(
  "district_total_ppl_map.png",
  fig6,
  width = 8,
  height = 8,
  dpi = 300
)
```

## Key results

```{r}
top10
```

```{r}
highest_districts <- district_total %>%
  slice_max(
    total_dispensings,
    n = 5
  )

highest_districts
```

```{r}
lowest_districts <- district_total %>%
  slice_min(
    total_dispensings,
    n = 5
  )

lowest_districts
```

## Change from 2021 to 2025

```{r}
change_top10 <- trend_top10 %>%
  group_by(
    TherapeuticGrp2
  ) %>%
  summarise(
    dispensing_2021 =
      total_dispensings[
        YearDisp == 2021
      ],
    dispensing_2025 =
      total_dispensings[
        YearDisp == 2025
      ],
    change =
      dispensing_2025 -
      dispensing_2021,
    percent_change =
      (
        dispensing_2025 -
          dispensing_2021
      ) /
      dispensing_2021 *
      100,
    .groups = "drop"
  ) %>%
  arrange(
    desc(change)
  )

change_top10
```

## Highest District + Therapeutic Group combinations

```{r}
highest_combinations <- district_top10 %>%
  arrange(
    desc(total_dispensings)
  ) %>%
  slice_head(
    n = 10
  )

highest_combinations
```

## Lowest District + Therapeutic Group combinations

```{r}
lowest_combinations <- district_top10 %>%
  arrange(
    total_dispensings
  ) %>%
  slice_head(
    n = 10
  )

lowest_combinations
```
