---
title: "R Notebook"
output: html_notebook
---

## Group 19 (so cool, the best)

We say most common dispensed pharmaceuticals - we are using NumPpl, how many people have received a medication, to track this, rather than the number of dispensings.

```{r}
library(dplyr)
library(ggplot2)
```

### Pharmaceuticals Lookup Metadata

Can combine to also look at TG1, or all three therapeutic groups together.

```{r}
metadata <- read.csv("PharmaceuticalsLookup.csv")
head(metadata)
```

### New Zealand Totals by TG2

```{r}
# New Zealand Totals by TG2
tg2 <- read.csv("Data_ByTG2.csv")

tg2 <- tg2 %>%
  filter(NumDisps != "<6", NumPpl != "<6") %>%
  mutate(
    NumDisps = as.integer(NumDisps),
    NumPpl = as.integer(NumPpl)
  )

counts <- tg2 %>%
  group_by(TherapeuticGrp2) %>%
  summarise(
    NumPpl_count = sum(NumPpl),
    .groups = "drop"
  )

print(counts, n = 123)
```

```{r}
# Arrange in Descending Order to Find Most Common 
counts <- counts %>% arrange(desc(NumPpl_count))
print(counts, n=123)
```

```{r}
countsAscending <- counts %>% arrange(NumPpl_count)

ggplot(countsAscending, aes(x = seq_along(NumPpl_count), y = NumPpl_count)) + 
  geom_line(color = "blue", linewidth = 1) +
  geom_point(color = "red", size = 2) +
  theme_minimal()
```

### Merge if you wanna compare all 3

```{r}
# Merged database: PharmaceuticalsLookup and Data_ByTG2

```

### Census Data (Not Yet Cleaned)

Note the census counts Hutt Valley and C&C as one DHB, must be combined in TG2 if wanting to merge.

```{r}
# NZ Census Data: Using 2023 to represent the post-covid trends captured by the Pharmac dataset, and is centre point of temporal data (2021-2025)
# This is a wide dataset

census2023 <- read.csv("census2023.csv")
```
